---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ggragged

<!-- badges: start -->
<!-- badges: end -->

ggragged extends the faceting system in [ggplot2](https://ggplot2.tidyverse.org/)
to ragged grids -- a hybrid between `facet_wrap()` and `facet_grid()`.

- `facet_ragged_rows()` groups panels into rows of (potentially) varying lengths.
- `facet_ragged_cols()` groups panels into columns of (potentially) varying lengths.

## Installation

You can install the development version of ggragged like so:

``` r
remotes::install_github("mikmart/ggragged")
```

## Example

Ragged grids can be useful, for example, to clearly separate nested
hierarchies in the panel layout:

```{r example}
library(ggplot2)
library(ggragged)

p <- ggplot(Indometh, aes(time, conc)) + geom_line()

# Panels for each subject, with cohorts on separate rows
p + facet_ragged_rows(
 vars(Cohort = 1 + Subject %in% 3:6),
 vars(Subject = as.character(Subject)),
 labeller = label_both
)
# Panels for each subject, with cohorts in separate columns
p + facet_ragged_cols(
 vars(Subject = as.character(Subject)),
 vars(Cohort = 1 + Subject %in% 3:6),
 labeller = label_both
)
```
